\toolname: file-based localization of numerical perturbations in data analysis pipelines

### 2020-04-08 paper.tex
finished first pass

We did not observe any within-OS difference in PreFreeSurfer. In
FreeSurfer, we identified 2 processes leading to within-OS differences due
to the use of pseudo-random numbers: image registration with
\texttt{mri\_segreg}, and cortical surface curvature estimations with
\texttt{mris\_curvature}. Fixing the random seed used in FreeSurfer removed
these differences.

@@

We identified four types of subjects with different PreFreeSurfer
provenance graphs (Table~\ref{table:data-clusters}). Differences between
subject types are coming from different  numbers of T1 and T2 images in the
raw data. We verified that
the provenance graphs were identical for all subjects of the same type, for
both versions of CentOS.

Figure~\ref{fig:pfs_heatmap} shows the frequency of non-reproducible pipeline processes
in PreFreeSurfer. 
Differences were observed in linear registration 
with FSL \flirt (in ACPC-Alignment, Brain Extraction, Distortion Correction, and
Atlas Registration), in non-linear registration with FSL \fnirt (in Brain Extraction 
and Atlas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL \texttt{maths}  (in Anatomical Average). 
Figure~\ref{fig:complete_pfs} shows a complete PreFreeSurfer labeled DAG, localizing 
the observed differences in the entire pipeline, for a given subject.
\begin{figure*}
\begin{minipage}{\linewidth}
\centering
  \includegraphics[width=\textwidth]{./figures/heatmap.pdf}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps (N=20). 
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  number of subjects where the corresponding process appears. For example,
  the \flirt tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 19 subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  17 subjects.
        }
  \label{fig:pfs_heatmap}
\end{minipage}

@@

To illustrate the magnitude of the differences,
Figure~\ref{fig:fnirt_result} compares \fnirt results in Brain Extraction
for a particular subject. Differences appear to be very substantial, in
particular in the frontal and parietal lobes. 
\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{figures/t2w_alignment.png} 
    \caption{Differences between T2 \fnirt results in PreFreeSurfer's Brain Extraction (CentOS6 vs 
    CentOS7). An animated version of the comparison is available \href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/pfs_t2w_alignment.gif}{here}.}
    \label{fig:fnirt_result}
\end{figure}

@@

However, FreeSurfer results still differ between conditions, due to the
propagation and amplification of differences created in PreFreeSurfer. We
observed the effect of this propagation in FreeSurfer results, as shown in
Figure~\ref{fig:tissue_class} for whole-brain segmentations. The
Dice coefficients associated with the 44 regions segmented by
FreeSurfer are shown in Figure~\ref{fig:scatter_plot}, showing that many 
structures have Dice coefficients below 0.9 for some subjects.
\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{figures/brain_segmentation_mni.png} 
  \caption{Sum of binarized differences between whole-brain FreeSurfer
  segmentations obtained from PreFreeSurfer processings in CentOS6 vs CentOS7
   (N=20). Segmentations were resampled and overlaid to the MNI152 volume
  template. An animated comparison of segmentations obtained for a particular subject is available
\href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/fs_brain_segmentation.gif}
{here}.
} 
  \label{fig:tissue_class}
\end{figure}

\begin{figure}
  \hspace*{-1cm}
  \includegraphics[width=1.2\linewidth]{figures/dice_regions.pdf}
    \caption{Dice coefficients between regions segmented by FreeSurfer in CentOS6 vs CentOS7 (N=20).}
    \label{fig:scatter_plot}
  \end{figure}

@@

\subsection{Key findings}

Linear and non-linear registration with FSL were found to
frequently lead to differences between results obtained with different
operating systems. This does not come as a surprise given the instabilities
associated with these processes. It also corroborates our previous findings
in~\cite{Glatard2015}, where fMRI pre-processing with FSL was found to vary across operating systems 
starting from the motion
correction step, a step that uses FSL's \flirt tool internally. It
would be relevant to investigate if the observed instability of
registration processes generalizes to other toolkits, or if it remains specific
to FSL. In view of the effect of small data perturbations in a variety of
toolboxes and processes, such as cortical surface extraction using
FreeSurfer and CIVET~\cite{Lewis2017-ll} or connectome estimation using
Dipy~\cite{kiar2019comparing}, it is probable that this observation
generalizes widely across toolboxes and requires a deeper investigation of
the stability of linear and non-linear registration.

While only a handful or processes were found non-reproducible across the
tested operating systems, the effect of such instabilities were found to
propagate widely in the pipelines, and to substantially impact FreeSurfer's
segmentations. This illustrates the need to conduct reproducibility studies
on entire pipelines rather than isolated processes. It also highlights the need 
for a deeper stability analysis of pipeline processes.

Furthermore, as shown in Figure~\ref{fig:pfs_heatmap}, the reproducibility of
a given tool may vary across subjects and across processing parameters. For
instance, linear registration with \flirt seems to be fully reproducible in
the Anatomical Average sub-pipeline, while it is highly non-reproducible in
ACPC Alignment. In Brain Extraction, the same tool was found reproducible
for some subjects only. Therefore, reproducibility studies need to be
performed on several subjects. While this is common practice in neurosciences, 
software tests are often executed only on a single dataset, to reduce the
associated computational load. Our results show that pipeline tests should 
encompass enough subjects to cover execution paths adequately.

@@

To evaluate the reproducibility of a pipeline, \toolname needs to execute
it 3 times in order to (1) record a first \reprozip trace, (2) save
transient files, and (3) compare results in the second condition. This is
one more execution than the theoretical minimum of 2. It might be possible
to further reduce this overhead, by executing at step (2) only the
processes depending on transient files.

@@

Future work will investigate in more details the numerical stability of
registration algorithms. Besides, we plan on using Monte-Carlo arithmetic
to inject controlled amounts of noise in the pipelines, and monitory
uncertainty propagation and amplification in the pipelines.

>>>>

We did not observe any within-OS difference in PreFreeSurfer. In
FreeSurfer, we identified 2 processes leading to within-OS differences due
to the use of pseudo-random numbers: image registration with
\texttt{mri\_segreg}, and cortical surface curvature estimations with
\texttt{mris\_curvature}. Fixing the random seed used in FreeSurfer removed
these differences. \note{Greg}{can you make some statement on the magnitude of these differences?}

@@

We identified four types of subjects with different PreFreeSurfer
provenance graphs (Table~\ref{table:data-clusters}). Differences between
subject types came from different numbers of T1 and T2 images in the
raw data. We verified that
the provenance graphs were identical for all subjects of the same type, for
both versions of CentOS.

Figure~\ref{fig:pfs_heatmap} shows the frequency of non-reproducible pipeline processes
in PreFreeSurfer. 
The processes identified as non-reproducible were observed in linear registration 
with FSL \flirt (in ACPC-Alignment, Brain Extraction, Distortion Correction, and
Atlas Registration), in non-linear registration with FSL \fnirt (in Brain Extraction 
and Atlas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL \texttt{maths}  (in Anatomical Average). 
Figure~\ref{fig:complete_pfs} shows a complete PreFreeSurfer labeled DAG, localizing 
the observed differences in the entire pipeline, for a given subject.
\begin{figure*}
\begin{minipage}{\linewidth}
\centering
  \includegraphics[width=\textwidth]{./figures/heatmap.pdf}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps (N=20). 
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  number of subjects where the corresponding process appears. For example,
  the \flirt tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 19 subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  17 subjects. \note{Greg}{colorbar doesn't match the numbers: it's from 0-1 but the numbers are 
  subject counts. I don't really have a preference on which, but harmonizing them to be the same unit should happen.}
        }
  \label{fig:pfs_heatmap}
\end{minipage}

@@

To illustrate the magnitude of the differences,
Figure~\ref{fig:fnirt_result} compares \fnirt results in Brain Extraction
for a particular subject. Differences appear to be very substantial, in
particular in the frontal and parietal lobes.\note{Greg}{This is very hard to see from the figure because the
luminosity of the two colormaps are so different - the contrast moving from orange to grayscale is difficult. Perhaps use two colors, green and purple for instance, with the same intensity to see if this is more easily seen?} 
\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{figures/t2w_alignment.png} 
    \caption{Differences between T2 \fnirt results in PreFreeSurfer's Brain Extraction (CentOS6 vs 
    CentOS7). An animated version of the comparison is available \href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/pfs_t2w_alignment.gif}{here}.}
    \label{fig:fnirt_result}
\end{figure}

@@

However, FreeSurfer results still differ between conditions, due to the
propagation and amplification of differences created in PreFreeSurfer. We
observed the effect of this propagation in FreeSurfer results, as shown in
Figure~\ref{fig:tissue_class} for whole-brain segmentations. The
Dice coefficients associated with the 44 regions segmented by
FreeSurfer are shown in Figure~\ref{fig:scatter_plot}, showing that many 
structures have Dice coefficients below 0.9 for some subjects. \note{Greg}{awkward "many" followed by "some". reword?}
\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{figures/brain_segmentation_mni.png} 
  \caption{Sum of binarized differences between whole-brain FreeSurfer
  segmentations obtained from PreFreeSurfer processings in CentOS6 vs CentOS7
   (N=20). Segmentations were resampled and overlaid to the MNI152 volume
  template. An animated comparison of segmentations obtained for a particular subject is available
\href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/fs_brain_segmentation.gif}
{here}.
} 
  \label{fig:tissue_class}
\end{figure}

\begin{figure}
  \hspace*{-1cm}
  \includegraphics[width=1.2\linewidth]{figures/dice_regions.pdf}
    \caption{Dice coefficients between regions segmented by FreeSurfer in CentOS6 vs CentOS7 (N=20).
\note{Greg}{I don't think the legend is important here. Focus on having values legible and removing those which
don't really illustrate the "point".}}
    \label{fig:scatter_plot}
  \end{figure}

@@

\subsection{Key findings}
\note{Greg}{feel free to ignore, but a peeve of mine is one heading directly following another. Is there some
kind of high-level summary you could put above this, or just omit the key-findings label?}
Linear and non-linear registration with FSL were found to
frequently lead to differences between results obtained with different
operating systems. This does not come as a surprise given the instabilities
associated with these processes. It also corroborates our previous findings
in~\cite{Glatard2015}, where fMRI pre-processing with FSL was found to vary across operating systems 
starting from the motion
correction step, a step that uses FSL's \flirt tool internally. It
would be relevant to investigate if the observed instability of
registration processes generalizes to other toolkits, or if it remains specific
to FSL. In view of the effect of small data perturbations in a variety of
toolboxes and processes, such as cortical surface extraction using
FreeSurfer and CIVET~\cite{Lewis2017-ll} or connectome estimation using
Dipy~\cite{kiar2019comparing}, it is probable that this observation
generalizes widely across toolboxes and requires a deeper investigation of
the stability of linear and non-linear registration.

While only a handful or processes were found non-reproducible across the
tested operating systems, the effect of such instabilities were found to
propagate widely in the pipelines, and to substantially impact the segmentations
created by FreeSurfer. This illustrates the need to conduct reproducibility studies
on entire pipelines rather than isolated processes. It also highlights the need 
for a deeper stability analysis of pipeline processes.

As is shown in Figure~\ref{fig:pfs_heatmap}, the reproducibility of
a given tool may vary across subjects and across processing parameters. For
instance, linear registration with \flirt seems to be fully reproducible in
the Anatomical Average sub-pipeline, while it is highly non-reproducible in
ACPC Alignment. In Brain Extraction, the same tool was found reproducible
for some subjects only. Therefore, reproducibility studies need to be
performed on several subjects. While this is common practice to some extent in neuroimaging, 
software tests are often executed only on a single dataset to reduce the
associated computational load. Our results show that pipeline tests should 
encompass enough subjects to cover execution paths adequately.

@@

To evaluate the reproducibility of a pipeline, \toolname needs to execute
it 3 times in order to (1) record a first \reprozip trace, (2) save
transient files, and (3) compare results in the second condition. This is
one more execution than the theoretical minimum of 2. It might be possible
to further reduce this overhead by executing at step (2) only the
processes depending on transient files.

@@

Future work will investigate in more details the numerical stability of
registration algorithms. Additionally, we plan on using Monte-Carlo arithmetic
to inject controlled amounts of noise in pipelines and monitor
uncertainty propagation and amplification in their results.
